All just Greedy for Money? - A prediction and Analysis of the interplay between Graduate Statistics and Salary Prospects in the STEM Field

### 2024-01-28 doc/DataLiteracyStudentProject/paper.tex
update

Our central contribution is a Vector autoregressive model (VAR) discussed in \Cref{sec:methods:var} that is built using just mentioned data. Combined with a
Partial autocorrelation function (PACF) the optimal value for lag
is determined, see \Cref{sec:methods:pacf}, to ensure that relevant structure in the data is captured by the model.
After verifying that the VAR model functions as expected and testing the model on a test set, \Cref{sec:methods} \textcolor{red}{subsection}, we
can predict all three input time series. As we are only interested in the number of students enrolled and the development of the salary, our research is 
focused on these two instances.
This allows us to conclude that salary has a significant impact on the amount of STEM students. Further, we can model different salary development scenarios, 
and see how the students would likely react (\Cref{sec:results}).

@@

\subsection{PACF}\label{sec:methods:pacf}
To find the optimal hyperparameter we used two well-established methods, the PACF and information-based criteria. Particularly the AIC and 
BIC. The PACF only considers the impact of the previous data point $S_{n-1}$ on $S_{n}$ to build a $(1-\alpha)$ confidence interval.
We choose $\alpha = 5\%$ and get get a 95\% Confidence Interval 
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates \\
        \hline
        Prediciton 1 & 1.257\% & 1.271\% & 3.944\% \\
        \hline
        Prediciton 2 & 4.843\% & 0.119\% & 1.722\%
    \end{tabular}
    \caption{Difference in percent between the true value and predicted value for students, salary, and graduates. Lag set to 1.}
\end{table}
The percentage differences between the true values especially for the students are quite high. To get better results
the AIC and BIC are used to find a better lag for the hyperparameter $p$ \textcolor{red}{is $p$ = lag?}. The abbreviation AIC
stands for Akaike Information Criterion and tries to find an optimum between the complexity of the model and fitting the data.
This should prevent overfitting. 
BIC is short for Bayesian Information Criterion and penalizes more complex models.
Therefore BIC leads to less complicated models than the ones from AIC, which sets the two apart.
The optimal lag based on AIC is 6, while for BIC it is 1, which is consistent with the PACF.
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates \\
        \hline
        Prediciton 1 & 0.484\% & 0.419\% & 0.111\% \\
        \hline
        Prediciton 2 & 1.346\% & 0.306\% & 0.939\%
    \end{tabular}
    \caption{Percentage difference between the true value and predicted value for students, salary, and graduates. Lag set to 5.}
\end{table}

@@

As you can see there is a major improvement in the prediction for the students in comparison to lag order 1. We decided to use lag 5, because the main focus of our 
of our research is to predict the students based on salary and graduates.
Another quite important aspect in the correlation analysis is the Granger Causality Test which we did on the output. 
Granger Causal implies that one time series is the result of another shifted time series.
In our case, we try to find out if e.g. the salary time series is just a shifted version of the students or the graduates time series.
The Granger causality test does exactly this, it uses the time series $S_i$ and tries to insert data points from $S_j$ or $S_k$ $i \neq j\neq k$
to predict the outcome of the model. After inserting the data points the test performs a t-test and a F-test, if both of these tests return statistically significant,
then we showed that $S_i$ could be expressed by a shifted input time series $S_j$ with $j\neq i$

>>>>

Our central contribution is a Vector autoregressive model (VAR) discussed in \Cref{sec:methods:var} that is built using just mentioned data. Combined with a
Partial autocorrelation function (PACF) the optimal value for lag
is determined, see \Cref{sec:methods:lag}, to ensure that relevant structure in the data is captured by the model.
After verifying that the VAR model functions as expected and testing the model on a test set, \Cref{sec:methods} \textcolor{red}{subsection}, we
can predict all three input time series. As we are only interested in the number of students enrolled and the development of the salary, our research is 
focused on these two instances.
This allows us to conclude that salary has a significant impact on the amount of STEM students. Further, we can model different salary development scenarios, 
and see how the students would likely react (\Cref{sec:results}).

@@

\subsection{Choosing lag}\label{sec:methods:lag}
To find the optimal hyperparameter we used two well-established methods, the PACF and information-based criteria. Particularly the AIC and 
BIC. The PACF only considers the impact of the previous data point $S_{n-1}$ on $S_{n}$ to build a $(1-\alpha)$ confidence interval.
We choose $\alpha = 5\%$ and get get a 95\% Confidence Interval 
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates \\
        \hline
        Prediciton 1 & 1.257\% & 1.271\% & 3.944\% \\
        \hline
        Prediciton 2 & 4.843\% & 0.119\% & 1.722\%
    \end{tabular}
    \caption{Difference in percent between the true value and predicted value for students, salary, and graduates. Lag set to 1.}
\end{table}
The percentage differences between the true values especially for the students are quite high. To get better results
the AIC and BIC are used to find a better lag for the hyperparameter $p$ \textcolor{red}{is $p$ = lag?}. The abbreviation AIC
stands for Akaike Information Criterion and tries to find an optimum between the complexity of the model and fitting the data.
This should prevent overfitting. 
BIC is short for Bayesian Information Criterion and penalizes more complex models.
Therefore BIC leads to less complicated models than the ones from AIC, which sets the two apart.
The optimal lag based on AIC is 6, while for BIC it is 1, which is consistent with the PACF.
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates \\
        \hline
        Prediciton 1 & 0.484\% & 0.419\% & 0.111\% \\
        \hline
        Prediciton 2 & 1.346\% & 0.306\% & 0.939\%
    \end{tabular}
    \caption{Percentage difference between the true value and predicted value for students, salary, and graduates. Lag set to 5.}
\end{table}

The percentage errors show that there is a significant improvement in the prediction of student numbers compared to the lag order of 1.
Setting lag to 5 results in even better errors for the domains of our interest, predicting students.

\subsection{Test for causality}\label{sec:methods:causality}
To ensure that the model does not try to find structure in the data, where there is none, we applied the Granger causality test.
Granger causal implies that one time series is the result of another shifted time series.
In this case, we try to find out if e.g. the salary time series is just a shifted version of the student's or the graduate's time series.
The test does verifies this by using series $S_i$ and trying to insert data points from $S_j$ or $S_k$ $i \neq j\neq k$
to predict the outcome of the model. After inserting the data points a t-test and a F-test are performed. If both
tests return statistical significance, then we showed that $S_i$ could be expressed by a shifted input time series $S_j$ with $j\neq i$.
\textcolor{red}{OK, and did we show that?}

@@
